\section{Introduction}
\label{sec:intro}

EEG foundation models (FMs) pretrained on large unlabelled corpora
have recently been introduced as a general-purpose substrate for
clinical electroencephalography
\citep{labram2024,cbramod2025,reve2025}. On downstream tasks with
well-established within-subject neural contrasts --- motor imagery,
event-related potentials, sleep staging --- reported performance has
been consistently strong
\citep{eegfmbench2025,adabrain2025,eegbench2025}. On small-$N$
resting-state EEG (rsEEG) cohorts, the regime most relevant to
psychiatric and neurodegenerative biomarkers, reported numbers
instead fragment widely across datasets, protocols, and backbones
\citep{brain4fms2026,kauffmann2025critique,smalln2025biorxiv}. A
single foundation-model architecture can appear to deliver
near-ceiling accuracy on one clinical cohort and chance-level
accuracy on another cohort of comparable size and recording quality,
with no accompanying explanation for the gap. This paper asks
\emph{why}.

Two complementary literatures bear on this question but neither
resolves it. Aggregate benchmarks enumerate downstream scores across
many tasks
\citep{brain4fms2026,eegfmbench2025,adabrain2025,eegbench2025},
confirming that fragmentation exists without explaining what controls
it. Protocol audits have identified subject-identity leakage under
trial-level cross-validation as a major inflation source
\citep{brookshire2024leakage,smalln2025biorxiv,kauffmann2025critique},
and subject-disjoint evaluation is now increasingly adopted. Both
developments address \emph{that} a problem exists; neither provides a
structural account of \emph{which} datasets should be expected to
fragment in \emph{which} directions. In the absence of such an
account, each new benchmark result on a small clinical cohort is
interpreted in isolation, and positive findings on one cohort are
implicitly extrapolated to cohorts with different labelling
paradigms.

We observe that FM downstream behaviour on small-$N$ clinical rsEEG
varies along two orthogonal axes specifiable from a dataset's
metadata and a single linear-probe measurement, before any model is
fine-tuned. Axis A is the \emph{subject relation} of the label:
whether the label admits a within-subject contrast (e.g., rest vs.\
mental arithmetic, or rested vs.\ sleep-deprived) or is defined as a
single scalar per subject (e.g., an Alzheimer's vs.\ control
diagnosis, or a trait stress score). Axis B is an \emph{observational}
report: whether a frozen linear probe on FM features separates the
label above chance under subject-disjoint cross-validation,
consistently across multiple backbones. We call the two outcomes
\emph{separable} and \emph{null-indistinguishable}, and emphasise
that this is a report of what the probe measures rather than an
imposed label-signal category. Crossing the two binary axes yields a
$2\times 2$ factorial of operating regimes, and we populate this
factorial with four small-$N$ clinical rsEEG datasets, one per cell.
The resting-state Stress cohort \citep{komarov2020stress} instantiates
the between-subject $\times$ null-indistinguishable corner; prior FM
evaluation on this cohort \citep{wang2025stress} provides a published
FM reference point that no other dataset in our factorial offers at
the same corner. The Alzheimer's--Frontotemporal Dementia cohort
ADFTD \citep{adftd} instantiates between $\times$ separable, EEGMAT
\citep{eegmat} instantiates within $\times$ separable, and a public
sleep-deprivation rsEEG cohort \citep{xiang2025sleepdep} instantiates
within $\times$ null-indistinguishable.

We make three contributions. \textbf{First}, to our knowledge this
is the first structured factorial comparison of EEG-FM behaviour
across (subject-labelling $\times$ linear-probe separability) axes,
with each cell populated by a matched small-$N$ clinical rsEEG
cohort. \textbf{Second}, we pair four diagnostics --- variance
decomposition, permutation null, within-subject direction
consistency, and causal anchor ablation via
FOOOF\footnote{``fitting oscillations and one-over-f'', the
parameterisation of aperiodic $1/f$ background vs.\ periodic peaks.}
and band-stop filtering --- with explicit scope conditions stating
the cell types in which each runs. The individual tools are
established in prior work; the contribution is the pairing of tool
and scope condition. Within-subject direction consistency runs only
in within-subject cells by construction, causal anchor ablation
returns an \texttt{absent} / $\alpha$-broadband / $1/f$-aperiodic
verdict per cell, and variance decomposition returns a
label-dominated vs.\ subject-dominated reading for any frozen FM
feature set. \textbf{Third}, we distil a per-cell mechanism account
and a four-gate pre-flight checklist that a practitioner can apply
before committing fine-tuning compute to a new clinical cohort.

We state the scope of our claims up front. With $n=1$ dataset per
cell, this paper is deliberately \emph{descriptive} rather than
predictive: we do not claim that the $2\times 2$ axes forecast
outcomes for unseen datasets, and the factorial is a way of
organising our own observations rather than a map we ask readers to
consult for new cohorts. What we do ask readers to take away is the
toolkit and its scope conditions: each diagnostic has a defined
input, output, and scope condition computable from EEG alone.
Sec.~\ref{sec:related} reviews EEG-FM benchmarks, subject-leakage
protocol audits, and prior taxonomies. Sec.~\ref{sec:methods}
details the four datasets, per-FM input normalisation,
subject-disjoint evaluation, and the method-level specification of
the four diagnostics. Sec.~\ref{sec:results} opens with a benchmark
landscape (Tab.~\ref{tab:master_performance}) situating each cell's
LP/FT balanced accuracy before the diagnostic analysis, then reports
the four diagnostics in turn. Sec.~\ref{sec:discussion} synthesises a
per-cell mechanism account and the pre-flight checklist, and
Sec.~\ref{sec:limitations} discusses scope.
