\section{Methods}
\label{sec:methods}

We describe the four-dataset factorial, the feature extraction and
evaluation protocol shared across cells, the fine-tuning recipe, and
the method-level specification of the four diagnostics whose per-cell
outputs are reported in Sec.~\ref{sec:results}.

\subsection{The $2\times 2$ factorial and cell assignment rationale}
\label{sec:methods_2x2}

\paragraph{Axis A --- subject relation of the label.}
Axis A is a property of the labelling paradigm, readable from dataset
metadata alone. A dataset is \emph{within-subject paired} when each
subject contributes recordings under both label classes and the
contrast can be evaluated inside a single subject (rest vs.\ arithmetic
in EEGMAT; rested vs.\ sleep-deprived in the SleepDep cohort). A
dataset is \emph{between-subject} when the label is defined as a
scalar per subject, either by trait (AD/FTD vs.\ HC in ADFTD) or by a
per-recording score that is empirically subject-stable. The
resting-state Stress cohort is labelled per recording by a DASS-21
threshold, but 14/17 subjects receive the same label across all of
their recordings; the remaining 3/17 straddle both classes with too
few within-subject contrasts to support within-subject decoding, so we
treat the Stress cell as between-subject throughout.

\paragraph{Axis B --- linear-probe separability under subject-disjoint
CV.} Axis B is an \emph{observational} property: we report whether a
frozen-feature linear probe separates the label above chance on a
dataset, not whether there ``is signal'' in the EEG. A cell is
\emph{separable} when at least two of the three foundation-model
backbones (Sec.~\ref{sec:methods_fm}) attain balanced accuracy above
chance with permutation-null $p<0.05$ under the primary evaluation
protocol (Sec.~\ref{sec:methods_eval}); otherwise the cell is
\emph{null-indistinguishable}. This definition deliberately uses
cross-FM consistency rather than any single backbone's result, so that
a single favourable architecture cannot pull a cell across the
threshold.

\paragraph{Cell assignments.} Each of the four cells is populated by
one publicly evaluable small-$N$ clinical rsEEG dataset. EEGMAT
(within $\times$ separable) and SleepDep (within $\times$
null-indistinguishable) share the within-subject paired paradigm but
differ in whether any FM's linear probe separates the label; ADFTD
(between $\times$ separable) and Stress (between $\times$
null-indistinguishable) share the between-subject labelling paradigm
but differ in the same way. The $2\times 2$ is descriptive by design:
with one dataset per cell we do not treat the factorial as a
predictive map, and state this scope boundary in
Sec.~\ref{sec:discussion}.

\subsection{Per-dataset specifications}
\label{sec:methods_datasets}

All four datasets are resting-state or rest-plus-task EEG cohorts of
comparable size (17--65 subjects) and recording quality (19--30
channels, 200\,Hz nominal sampling after per-dataset resampling).
Preprocessing is uniform across datasets: a $1$--$45$\,Hz zero-phase
Butterworth band-pass, DC removal, and $5$\,s non-overlapping epochs
(trainable overlap is set per dataset, Sec.~\ref{sec:methods_ft}).

\textbf{Stress-DASS.} 17 subjects contributing 70 resting-state
recordings with concurrent DASS-21 and DSS self-report;
\citet{komarov2020stress} describes the collection protocol and
\citet{wang2025stress} reports the previously published FM evaluation
on a subset of this cohort. 30-channel scalp montage at 1000\,Hz
resampled to 200\,Hz. Binary label from a DASS-21 stress threshold
per recording; regime: between $\times$ null-indistinguishable.

\textbf{ADFTD.} 65 subjects from the public Alzheimer's disease /
frontotemporal dementia scalp-EEG cohort \citep{adftd}. A single
resting-state recording per subject is split into three non-overlapping
5\,s segments to form 195 samples evaluated under subject-disjoint
cross-validation. 19-channel 10--20 montage at 500\,Hz resampled to
200\,Hz. Three-class label collapsed to binary (AD $\cup$ FTD vs.\ HC)
following the \citet{lead2024} convention; regime: between $\times$
separable.

\textbf{EEGMAT.} 36 subjects contributing 72 recordings under two
conditions (3-min eyes-closed rest and a 1-min serial-subtraction
arithmetic task) \citep{eegmat}, providing a within-subject paired
state contrast. 19-channel 10--20 montage at 500\,Hz resampled to
200\,Hz. Regime: within $\times$ separable.

\textbf{SleepDep.} 36 subjects contributing 72 recordings under two
conditions (normal sleep baseline vs.\ 24\,h sleep deprivation)
\citep{xiang2025sleepdep}, providing a second within-subject paired
contrast at a recording-pair topology matched to EEGMAT. 19-channel
montage resampled to 200\,Hz. Regime: within $\times$
null-indistinguishable.

\subsection{Foundation model feature extraction and per-FM input
normalisation} \label{sec:methods_fm}

We evaluate three open-weight EEG foundation models spanning
complementary pretraining objectives and tokenisation strategies:
LaBraM \citep{labram2024}, CBraMod \citep{cbramod2025}, and REVE
\citep{reve2025}. Each backbone receives a $5$\,s $\times C$-channel
input at 200\,Hz and emits a pooled embedding of dimension
$d \in \{200, 200, 512\}$. The pooling rule is fixed by the backbone
(class-token for LaBraM, temporal-mean for CBraMod, temporal-mean for
REVE) and we do not re-pool.

\paragraph{Per-FM input normalisation.} The three backbones require
different input scales, and treating these as interchangeable silently
destroys evaluation. LaBraM is fed per-window z-scored input, matching
its pretraining recipe. CBraMod and REVE are fed raw microvolt-scale
input: CBraMod internally divides by $100$ inside its patch embedding
(so z-scoring before the backbone double-scales the signal toward
zero), and REVE's linear patch embedding was trained on microvolt-scale
data and is scale-sensitive. We verify the per-FM norm against each
original repository before every experiment and hold it fixed across
cells; it is not swept as a hyperparameter. This asymmetry becomes
load-bearing in Sec.~\ref{sec:results} when we interpret the FOOOF
ablation signatures across models.

\paragraph{Feature extraction modes.} Frozen linear probe (LP) uses
the backbone in \texttt{eval} mode with no gradient; the pooled
embedding is passed to an $L_2$-penalised logistic regression with
\texttt{class\_weight=balanced} and per-window predictions are
aggregated to a recording-level decision by majority vote. Full
fine-tuning (FT) unfreezes the backbone and adds a two-layer head
trained end-to-end under the recipe in Sec.~\ref{sec:methods_ft}.
Architecture integrates through a single \texttt{BaseExtractor}
contract (\texttt{(B, C, T)} $\rightarrow$ \texttt{(B, d)}), so the
training loop is backbone-agnostic.

\subsection{Evaluation protocol}
\label{sec:methods_eval}

\paragraph{Primary protocol --- subject-disjoint 5-fold CV.} All
cells report balanced accuracy (BA) under
\texttt{StratifiedGroupKFold} with $K=5$ grouped by subject so that no
subject's recordings appear in both train and test folds of a given
split \citep{brookshire2024leakage}. Aggregation is recording-level:
per-window posteriors are averaged within a recording, thresholded at
0.5, and compared to the recording label. This is the single
evaluation rule used for the factorial axis B classification and the
benchmark landscape (Tab.~\ref{tab:master_performance}).

\paragraph{Secondary protocol --- within-subject LOO (within-subject
cells only).} For EEGMAT and SleepDep we additionally evaluate a
leave-one-subject-out within-subject probe that, for each held-out
subject, fits a probe on that subject's own recording pair and
reports its BA. This protocol isolates state signal from subject
structure and supports the direction-consistency diagnostic in
Sec.~\ref{sec:methods_wsci}. It is inapplicable to Stress and ADFTD
by label design.

\paragraph{Multi-seed requirement.} Small-$N$ cells exhibit run-to-run
variability of $\pm 5$--$10$\,pp even under
\texttt{cudnn.deterministic=True}. We require that every balanced
accuracy claim is averaged over at least three training seeds
($\{42, 123, 2024\}$) and report mean $\pm$ sample standard deviation.

\subsection{Fine-tuning recipe}
\label{sec:methods_ft}

FT uses AdamW with cosine-annealed learning rate, focal cross-entropy
loss (to absorb the mild class imbalance on Stress and ADFTD), label
smoothing disabled, layer-wise learning-rate decay set to $1.0$
(uniform), and $3$ epochs of warm-up with the encoder frozen for the
first warm-up epoch. SMA-$3$ smoothed early stopping with patience
$15$ is applied on the held-out fold's balanced accuracy. The head is
a two-layer MLP with hidden width $128$ and dropout $0$. This recipe
is applied uniformly across Stress, ADFTD, and EEGMAT with
\texttt{lr}~$=1\mathrm{e}{-5}$, weight decay $0.05$, and batch size
$32$; each foundation model keeps its native encoder learning-rate
scale (LaBraM and CBraMod $0.1$, REVE $0.1$ with a slightly higher
global \texttt{lr}~$=3\mathrm{e}{-5}$, matching each backbone's
downstream-tuning convention in its original repository).

\paragraph{Per-cell deviation on SleepDep.} SleepDep's effective trial
count per subject (two recordings per subject, each yielding a small
number of 5\,s epochs after epoching) forces a smaller batch
size~$= 4$ to keep gradient statistics well-conditioned, and we reduce
weight decay to $0.01$ and disable augmentation overlap for the same
reason. This is the only per-dataset deviation from the shared recipe,
applied uniformly across the three FMs for SleepDep and not tuned
per backbone. No per-dataset learning-rate sweep is performed in any
cell; the HP search performed for Stress (archived in
\texttt{results/hp\_sweep/}) is reported only in Appendix~C.

Permutation-null runs (Sec.~\ref{sec:methods_null}) use the same
recipe as the observed-BA runs within each cell, so that the null
and observed distributions are calibrated under matched HP.

\subsection{Diagnostic toolkit --- method specifications}
\label{sec:methods_toolkit}

The toolkit consists of four diagnostics, each defined by an input, an
output, and a scope condition stating the cell types in which it
runs. Per-cell results are reported in Sec.~\ref{sec:results}; this
subsection gives the computational specification.

\subsubsection{Variance decomposition of frozen representations}
\label{sec:methods_variance}

For each dataset $\times$ FM cell, we compute pooled per-window frozen
features and decompose their variance across three orthogonal
directions: a \emph{label} axis (the linear discriminant solving
the recording-level task), a \emph{subject} axis (the top principal
component of the subject-centred feature matrix), and a
\emph{residual} axis (everything else, i.e.\ the leftover trace after
projecting onto the label and subject subspaces). We report
\texttt{label\_frac}, \texttt{subject\_frac}, and
\texttt{residual\_frac} as fractions of total feature variance,
estimated by a cluster bootstrap over subjects ($B=2{,}000$)
\citep{field2007clusterbootstrap} so that the confidence interval
respects the nested structure of recordings within subjects.
\emph{Scope condition:} this diagnostic runs in all cells regardless
of axis A or axis B.

\subsubsection{Permutation-null test}
\label{sec:methods_null}

For each cell and each FM we construct a null distribution by
retraining the full pipeline on label-shuffled data. The shuffle level
is chosen to match the labelling paradigm of the cell. For
between-subject cells (ADFTD: subject-level shuffle; Stress:
recording-level shuffle, since the labelling is per recording), we
permute the label vector over the appropriate unit and refit
end-to-end. For within-subject cells (EEGMAT, SleepDep) we permute the
recording-level label within the pair-level block. Each cell runs
30 null seeds under the same recipe as the observed run, and a
left-tail $p$-value is reported as the fraction of null BAs meeting or
exceeding the observed mean. \emph{Scope condition:} all cells; the
shuffle level is prescribed by the cell's labelling paradigm.

\subsubsection{Within-subject direction consistency}
\label{sec:methods_wsci}

For within-subject paired cells, we measure whether frozen FM features
represent the per-subject state contrast in a consistent direction
across subjects. For each subject $s$ with two recordings labelled
$y_s \in \{0,1\}$, we compute a per-subject contrast vector
$\Delta_s = \bar f_{s,1} - \bar f_{s,0}$ in feature space, where
$\bar f_{s,y}$ is the window-mean embedding for that recording. The
within-subject direction consistency index (WSCI) is the mean pairwise
cosine similarity over $\{\Delta_s\}_s$; a high WSCI implies subjects
encode the state transition along a shared axis, whereas WSCI near
zero implies idiosyncratic directions. We additionally report the
leave-one-subject-out BA of a linear probe trained on the within-subject
contrasts, as a decoding complement to the geometric WSCI.
\emph{Scope condition:} within-subject paired cells only (EEGMAT,
SleepDep); not applicable to Stress and ADFTD by label design, and we
mark the diagnostic as \emph{inapplicable} in those cells rather than
imputing a value.

\subsubsection{Causal anchor ablation (FOOOF + band-stop)}
\label{sec:methods_anchor}

We intervene on the EEG at two complementary spectral levels and
observe how the FM's frozen probe responds.

\emph{(i) FOOOF aperiodic / periodic ablation.} For each recording we
fit the FOOOF decomposition \citep{donoghue2020fooof} per channel on
the $1$--$45$\,Hz band, yielding an aperiodic $1/f$ component
(parameterised by offset and slope) and a set of periodic peaks. We
construct three phase-preserving reconstructions: (a)
\emph{aperiodic-removed}, replacing the $1/f$ background with a flat
spectrum while preserving the periodic peaks and the original phase
spectrum; (b) \emph{periodic-removed}, zeroing the peak components
while preserving the aperiodic background; and (c) \emph{both-removed},
a joint-ablation control. Each reconstruction is re-extracted through
the FM and probed with a subject-level one-vs-rest probe
(``subject-ID probe'') and a subject-disjoint state probe, each
averaged over 8 seeds. A drop in the state probe under
aperiodic-removal together with near-zero drop under periodic-removal
yields a \emph{$1/f$-aperiodic} anchor verdict; a drop under both
yields an \emph{$\alpha$-broadband} verdict (in-band peak plus
in-band $1/f$ tail jointly necessary); no drop in either yields an
\emph{absent-anchor} verdict.

\emph{(ii) Band-stop ablation.} As a second, model-free perturbation we
apply a $6$th-order Butterworth band-stop at canonical
$\delta$/$\theta$/$\alpha$/$\beta$ bands and re-extract features.
Cosine distance between the perturbed and original pooled embeddings
per channel quantifies which band each FM attends to; the panel is
reported for interpretability rather than as the primary anchor
verdict.

\emph{Scope condition:} all cells. The intervention is computed from
raw EEG before any FM is trained, so the anchor verdict is in
principle computable without running a foundation model --- the
FM-side probe measures whether the backbone has learned to depend on
that anchor.
