\section{Related Work}
\label{sec:related}

\subsection{EEG Foundation Models and the cross-subject clinical
  challenge}
The past two years have seen the emergence of large EEG foundation
models pretrained on aggregated public corpora: LaBraM
\citep{labram2024} (vector-quantised neural tokenizer,
$\sim\!100$\,M parameters), CBraMod \citep{cbramod2025} (criss-cross
attention, $\sim\!100$\,M parameters), REVE \citep{reve2025} (linear
patch embedding, $\sim\!1.4$\,B parameters), and a growing family of
encoder variants surveyed in recent reviews. Each reports competitive
or state-of-the-art performance on selected downstream tasks ---
typically sleep staging, seizure detection, or event-related
potentials. Performance on \emph{cross-subject affective and clinical
state prediction} has been more mixed.

Four recent large-scale benchmarks have independently stress-tested FM
claims under strict cross-subject evaluation. Brain4FMs
\citep{brain4fms2026}, EEG-FM-Bench \citep{eegfmbench2025}, and
AdaBrain-Bench \citep{adabrain2025} each report settings where current
FMs fail to outperform classical pipelines or compact from-scratch
models. Most directly relevant is EEG-Bench \citep{eegbench2025},
which covers 14 datasets $\times$ 11 clinical tasks (epilepsy,
schizophrenia, Parkinson, OCD, TBI) under strict cross-subject splits
and concludes that ``while foundation models achieve strong
performance in certain settings, simpler models often remain
competitive, particularly under clinical distribution shifts.'' Our
work complements these benchmark audits by providing a
\emph{mechanistic} explanation for the observed pattern rather than a
further benchmark enumeration: we decompose the subject-versus-label
variance structure of FM representations, conduct a controlled paired
within-subject experiment that isolates contrast strength as the
operative variable, and confirm architecture-independence of the
ceiling on a single small clinical dataset where seven architectures
converge. Kauffmann et al.\ \citep{kauffmann2025critique} present the
most explicit position-paper-style critique of EEG FM progress to
date, noting that ``linear probing results were relatively worse''
than full fine-tuning and that ``general time-series foundation models
sometimes outperformed EEG-FMs'' --- observations consistent with our
finding that fine-tuning acts as \emph{projection} onto
subject-dominated axes rather than representation rewrite.

The cross-subject psychopathology decoding problem has also been
explicitly elevated to competition-track status by the EEG Foundation
Challenge 2025 \citep{eegchallenge2025}, whose explicit framing
positions cross-subject + clinical psychopathology prediction as
\emph{the} open problem for this field. This motivates the clear
pre-benchmark diagnostic protocol our paper supplies.

\subsection{Subject leakage and subject-level cross-validation}
The systematic impact of subject leakage on EEG deep learning
performance has been documented for nearly a decade
\citep{roy2019review,brookshire2024leakage}. Brookshire et al.\
\citep{brookshire2024leakage} is the current canonical reference for
trial-level vs subject-level CV gap quantification in translational
EEG deep learning. Our Wang et al.\ \citep{wang2025stress}
re-analysis is, to our knowledge, the first independent subject-level
audit of a claimed ``EEG FM beats classical stress classification''
result; we find that a single protocol correction accounts for the
majority of the reported gap.

\subsection{Statistical power and small-$N$ reliability in EEG}
The reliability of small-$N$ deep learning findings on EEG has
recently been quantified at unprecedented scale. A multisite
psychiatry study of $N = 2{,}874$ resting-state EEG recordings
\citep{smalln2025biorxiv} concludes that ``results from small samples
were unstable, with inflated and highly variable effect sizes,''
converging only at large $N$. This provides the statistical-power
backdrop against which our UCSD Stress (70 rec / 14 positive) audit
operates: any claim at this cohort size is fragile by construction,
and our permutation-null analysis quantifies this fragility directly.

\subsection{Fine-tuning dynamics and cross-modality precedents}
Our observation that FM fine-tuning acts as \emph{projection} rather
than \emph{rewrite} on EEG downstream tasks has cross-modality
analogues in recent large-scale language and vision work. Chen et al.\
\citep{chen2025sft} (757 fine-tuned models $\times$ 10 base
architectures $\times$ 10 datasets) report that ``the inductive biases
of the base model outweigh the specific SFT corpus in determining the
final representation'' --- a cleaner LLM-scale instantiation of the
same phenomenon. In the vision domain \citep{li2025robustness}
document an analogous erosion pattern under fine-tuning. For EEG
emotion recognition, \citep{multidataset2025} proposes a cross-dataset
covariance-alignment loss that partially mitigates the inter-subject
variability problem by design; our contribution is complementary ---
we \emph{diagnose} the structural problem, whereas such methods
\emph{propose} solutions that future work may test against our SDL
diagnostic.

\subsection{Cognitive-load alpha desynchronisation}
The neurophysiological basis for distinguishing EEGMAT (rest vs
mental arithmetic) from UCSD Stress (DASS state) as targets for
within-subject FM evaluation rests on decades of oscillatory EEG
literature. Klimesch \citep{klimesch1999} established alpha
desynchronisation as the canonical cognitive-load signature;
Pfurtscheller and Aranibar \citep{pfurtscheller1979} provided the
event-related desynchronisation framework. EEGMAT rest vs arithmetic
produces a within-subject effect that is robust, spectrally localised
to alpha, and reliably induced across participants. By contrast, no
published EEG literature establishes a per-recording within-subject
correlate of DASS-21 or the Daily Stress Scale (DSS) at the resolution
used in the Wang et al.\ benchmark. This distinction --- independent
of any FM analysis --- is what our paired experiment operationalises.

\subsection{Subject biometric identity in EEG}
The dominance of subject-identifying information in EEG
representations is not a new observation in the biometric literature
\citep{campisi2014,marcel2007}. What is new in our subject-atlas
analysis is the quantification of subject-vs-label variance inside
\emph{pretrained FM} representations using variance decomposition and
cross-dataset RSA, providing a representation-level bridge between the
biometric literature and the downstream-task literature that has
largely overlooked it.

\subsection{Positioning of the present work}
Our contribution differs from three adjacent literatures in the
following specific ways. Against concurrent large-scale EEG-FM
benchmark audits \citep{eegbench2025,kauffmann2025critique,eegfmbench2025,adabrain2025},
which empirically document \emph{that} FMs sometimes fail to exceed
classical baselines on cross-subject clinical tasks, we supply the
mechanistic layer: a controlled within-subject paired experiment
that isolates neural contrast strength as the operative variable,
plus a pre-benchmark diagnostic protocol that predicts failure
before the benchmark is run. Against the EEG biometric literature
\citep{campisi2014,marcel2007}, which has long established
subject-discriminability as a property of classical EEG features, we
extend the quantification into pretrained FM representations and
connect it explicitly to downstream clinical classification outcomes.
Against the small-$N$ reliability critique
\citep{smalln2025biorxiv,varoquaux2017}, which treats statistical
instability as a general problem of effect-size reporting, we
identify a specific structural cause of the instability in the EEG-FM
setting (subject-dominance of the representation) and show that it
interacts with within-subject contrast strength to produce a
two-regime behaviour pattern. The result is a diagnosis rather than a
remediation; methods that explicitly target subject-dominance
during pretraining \citep{multidataset2025} are a complementary
direction that future work can test against the present diagnostic.
